### 2021-11-11 NSF2021/Description/specificaim2.tex
done up to 4.2.2

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. In Specific
Aim 2, we propose to develop efficient high-order numerical methods to
simulate dynamic assembly of particles under HAP potential in a viscous
solvent. As described in~\eqref{eq:stokes1}--\eqref{eq:stokes3} and
illustrated in Figure~\ref{fig:flow_map}, the solvent phase is modeled
by the Stokes equations for an incompressible, viscous fluid, and these
equations are coupled to the screened-Laplace equation~\eqref{eq:SL}
through viscous and hydrophobic stress balance.

%\label{subsec:specific_aim_2}
%\begin{wrapfigure}[14]{r}{0.61\textwidth}
%  \centering{\includegraphics[width=0.60\textwidth]{Figures/domain.pdf}}
%    \vspace{-15pt}
%    \caption{\label{fig:flow_map} \footnotesize A collection of 3D
%    amphiphilic particles suspended in a solvent. The solvent velocity
%    satisfies the Stokes equations, and the hydrophobic forces depend on
%    the solvent activity that satisfies the screened Laplace equation.
%    The geometry is updated by solving the mobility problem.} 
%\end{wrapfigure}
The HAP model requires solving exterior Dirichlet problems for the order
parameter~\eqref{eq:SL} and the Stokes
equations~\eqref{eq:stokes1}--\eqref{eq:stokes3}. If discretizing these
equations with stencil-based numerical methods, the computational domain
must be truncated, the fluid volume must be discretized, and artificial
boundary conditions must be imposed. The artificial boundary conditions
introduce additional error, and discretizing the volume results in an
excessively large system of equations. Moreover, it is difficult to
obtain high-order discretizations when the boundary surfaces are
irregular.

Alternatively, we will take advantage of the linearity of the Stokes
equations, and also apply a Newton's method to the HAP
model~\eqref{eq:SL}, where each iteration requires the solution of
linear elliptic PDEs. Then we can represent solutions of these elliptic
PDEs as layer potentials
\begin{align}
  \label{eq:LP}
  u(\mathbf{x}) = \mathcal{K}[\sigma](\mathbf{x}) = 
  \int_{\partial\Omega} K(\mathbf{x},\mathbf{y})
  \sigma(\mathbf{y})\,dS,
\end{align}
where $\sigma$ is an unknown density function, and
$K(\mathbf{x},\mathbf{y})$ is a derivative of the fundamental solution
of the PDE. By matching the prescribed boundary condition at
$\mathbf{x}_0 \in \partial\Omega$ with the limit of
equation~\eqref{eq:LP} as $\xx \rightarrow \xx_0$, a boundary integral
equation (BIE) for the density function is formed.

Numerical methods to solve BIEs have several advantages over their
PDE-based counterpart: by construction, the layer
potential~\eqref{eq:LP} satisfies far-field conditions; only
$\partial\Omega$ needs to be discretized; carefully chosen kernels
$K(\mathbf{x},\mathbf{y})$ result in a well-conditioned linear system
that is solved with a mesh-independent number of GMRES iterations; and
high-order or spectral accuracy is attainable with appropriate
quadrature methods. The PIs have demonstrated that BIEs are a powerful
tool to simulate two-dimensional suspensions of amphiphilic
particles~\cite{Fu2018_SIAM, FuQuRyYo20}. In addition to improving
two-dimensional simulations, the present proposal extends the results to
three dimensions using a three-dimensional BIE formulation of the
screened Laplace equation~\cite{ying_2006} and a well-conditioned BIE
formulation of the mobility problem~\cite{manasthesis, rac-gre2016}.

Significant challenges to solving BIEs include solving largely stem from
solving dense linear systems. First, \S\ref{subsec:AC} describes how the
non-linear Allen-Cahn equation~\eqref{eq:SL} will be iteratively solved
using BIE formulations. Then, \S\ref{subsec:NumericalIssues} describes
how we will combine preconditioning strategies with fast summation
methods to accelerate the time to solution, and
\S\ref{subsec:timeStepping} describes adaptive time stepping methods,

@@

In the PIs collaboration, they have developed numerical methods to
solve~\eqref{eq:SL} when $f(u) = u^2$. With this choice of $f(u)$, the
resulting PDE is the linear screened Laplace equation which is
efficiently solved with a double-layer potential representation. For a
more general $f$, such as a double well potential, we will interpret the
solution of~\eqref{eq:SL} as a root finding problem for the operator
\begin{align}
  \label{eq:F}
  J[u] = -\rho^2 \Delta u + f'(u).
\end{align}
This non-linear PDE will be solved with Newton's method
\begin{align*}
  u^{N+1} = u^{N} - \alpha (dJ^N)^{-1} F(u^N),
\end{align*}
where $dJ^N = -\rho^2 \Delta + f''(u^N)$ is the Fr\'{e}chet derivative
of $F$, and $\alpha$ is the step size which will be determined using
backtracking. As expected, the biggest computational challenge is
solving
\begin{align}
  \label{eq:NewtonUpdate}
  - \rho^2 \Delta v + f''(u^{N}) v = J[u^N].
\end{align}
\begin{wrapfigure}[]{l}{1.8in}
  \vspace{-46pt}
  \centering
  \input{figures/doublewell.tikz} 
  \caption{\label{fig:CA} (a) The double well potential $u_0 =
  0.5$, $u_1 = 2$, $a = 1$, $b = 0.4$. (b) The solution of the
  Allen-Cahn equation~\eqref{eq:SL}. (c) The convergence of the exact
  and inexact Newton's methods.}
\end{wrapfigure}
For a general double well potential $f(u)$, this is a variable
coefficient elliptic PDE, and developing an integral equation
formulation is challenging since the fundamental solution is not easily
determined. We propose to instead use an inexact Newton's method where
$dJ^{-1}$ is defined in terms of the piecewise quadratic approximation
\begin{align*}
  f(u) \approx \tilde{f}(u) = bu + \begin{cases}
    a(u - u_0)^2, &\mbox{if } u \leq (u_0 + u_1)/2 \\
    (u - u_1)^2, &\mbox{if } u > (u_0 + u_1)/2.
  \end{cases}
\end{align*}
In this case, the update step of Newton's method~\eqref{eq:NewtonUpdate}
amounts to solving the PDE
\begin{align}
  \label{eq:screenedPoisson}
  -\rho^2 \Delta v + \tilde{f}''(u^{N})v = J[u^N],
\end{align}
which is a screened Poisson equation with piecewise a constant screening
coefficient. We test the efficacy of exact and inexact Newton's method
applied to~\eqref{eq:F} by considering a finite difference method
applied to its one-dimensional version in $(0,1)$ with the Dirichlet
boundary conditions $u(0) = u_0$ and $u(1) = u_1$. These parameters
result in the double well potential illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the max norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
piecewise constant approximate (blue curve).

Having demonstrated that the inexact Newton method converges in one
dimension, it will be used to solve the Allen-Cahn equation in two more
or more dimensions. Since we are interested in solving~\eqref{eq:SL} in
general unbounded geometries, we will develop an integral equation
formulation. We first partition $\Omega$ into $\Omega_0 = \{\xx \in
\Omega \:|\: u(\xx) < (u_0 + u_1)/2\}$ and $\Omega_1 = \{\xx \in \Omega
\:|\: u(\xx) > (u_0 + u_1)/2\}$, and we let $\Lambda = \{\xx \:|\:
u(\xx) = (u_0 + u_1)/2 \}$ be their common boundary points. We then let
$G_0$ be the fundamental solution of $(-\rho^2 \Delta - 2a)$ and $G_1$
be the fundamental solution of $(-\rho^2 \Delta + 2)$. Then, the
solution of~\eqref{eq:screenedPoisson} can be written as $u = u^P + u^H$
where $u^P$ is a volume potential and $u^H$ is a layer potential. The
volume potential is
\begin{align*}
  u^{P}(\xx) = \int_{\Omega_i} G_i(\xx,\yy) J[u^N](\yy)\, d\yy, 
    \quad i = 1,2,
\end{align*}
but it is discontinuous across $\Lambda$ and does not satisfy the
prescribed hydrophobic boundary conditions on the particles $\Gamma_i$.
The boundary conditions are enforced with the layer potential
\begin{align*}
  u^{H}(\xx) = \begin{cases}
    \displaystyle\int_{\Lambda} G_0(\xx,\yy) \eta_0(\yy) \, ds_\yy + 
    \displaystyle\int_{\Gamma} \pderiv{}{\nn_\yy} G_0(\xx,\yy)
      \sigma(\yy) \,ds_\yy, \quad & \xx \in \Omega_0, \\
    \displaystyle\int_{\Lambda} G_0(\xx,\yy) \eta_1(\yy) \, ds_\yy,
     & \xx \in \Omega_1.
  \end{cases}
\end{align*}

Then, assuming that the particles are all contained in the
hydrophobic region, a solution of the screened Laplace equation
(i.e.~equation~\eqref{eq:screenedPoisson} when $G[u^N] = 0$) is
\begin{align*}
  u(\xx) = \frac{1}{2\pi} \int_{\bd\Omega} \pderiv{}{\nn_\yy} K_0 \left(
    \frac{|\xx - \yy|}{\rho}\right) \sigma(\yy) ds_\yy + 
    \frac{1}{2\pi} \int_{\Gamma} K_0 \left(
    \frac{|\xx - \yy|}{\rho} \right) \eta(\yy) ds_\yy,
\end{align*}

>>>>

\subsection{Specific Aim 2: Efficient, high-order methods for
large-scale simulations}
% ----------------------------------------------------------------------
Over the past decade, there has been an explosion of interest in
small-scale processes that utilize capillary forces and van der Waals
interactions to coordinate movement and bind microscopic components in
solvent~\cite{Pandey2011, Zhang2017, Siontorou2017}. Additionally,
hydrodynamic interactions are central to fabricate complex,
three-dimensional microstructures~\cite{Dasgupta2017, Leong2007,
Reynolds2019, Cho2010}. Hydrodynamic effects cannot be ignored since
they, like for the rates of biological functions like pore
dynamics~\cite{RYHAM20112929}, set the rate of dissipation. In Specific
Aim 2, we propose novel and efficient high-order numerical methods to
simulate dynamic assembly of particles under HAP potential in a viscous
solvent. As described in~\eqref{eq:stokes1}--\eqref{eq:stokes3} and
illustrated in Figure~\ref{fig:flow_map}, the solvent phase is modeled
by the Stokes equations for an incompressible, viscous fluid, and these
equations are coupled to the screened-Laplace equation~\eqref{eq:SL}
through viscous and hydrophobic stress balance.

%\label{subsec:specific_aim_2}
%\begin{wrapfigure}[14]{r}{0.61\textwidth}
%  \centering{\includegraphics[width=0.60\textwidth]{Figures/domain.pdf}}
%    \vspace{-15pt}
%    \caption{\label{fig:flow_map} \footnotesize A collection of 3D
%    amphiphilic particles suspended in a solvent. The solvent velocity
%    satisfies the Stokes equations, and the hydrophobic forces depend on
%    the solvent activity that satisfies the screened Laplace equation.
%    The geometry is updated by solving the mobility problem.} 
%\end{wrapfigure}
The HAP model requires solving exterior problems for the order
parameter~\eqref{eq:SL} and the Stokes
equations~\eqref{eq:stokes1}--\eqref{eq:stokes3}. We will develop
methods based on combinations of volume and layer potentials. Numerical
methods to solve these integral equations (IEs) have several advantages
over their PDE-based counterpart: by construction, they satisfy
far-field and mass conservation conditions; only $\partial\Omega$ needs
to be discretized for layer potentials; carefully chosen IEs result in a
well-conditioned linear system that is solved with a mesh-independent
number of GMRES iterations; and high-order or spectral accuracy is
attainable with appropriate quadrature methods. The PIs have
demonstrated that IEs are a powerful tool to simulate two-dimensional
suspensions of amphiphilic particles~\cite{Fu2018_SIAM, FuQuRyYo20}.
Moreover, elements of these formulations have been described in three
dimensions~\cite{ying_2006, manasthesis, rac-gre2016}. This proposal
will extend the two-dimensional methods to more general amphiphilic
particle models that satisfy~\eqref{eq:SL}, and all the methods can
naturally be extended to three dimensions. \S\ref{subsec:AC} describes
how the non-linear Allen-Cahn equation~\eqref{eq:SL} will be iteratively
solved using BIE formulations. Then, \S\ref{subsec:NumericalIssues}
describes how we will combine preconditioning strategies with fast
summation methods to accelerate the time to solution, and
\S\ref{subsec:timeStepping} describes adaptive time stepping methods,

@@

The PIs have developed numerical methods to solve~\eqref{eq:SL} when
$f(u) = u^2$---this choice results in the linear screened Laplace
equation. For a more general $f$, such as a double well potential, we
will interpret the solution of~\eqref{eq:SL} as the root finding problem
\begin{align}
  \label{eq:F}
  J[u] = -\rho^2 \Delta u + f'(u).
\end{align}
This non-linear PDE will be solved with Newton's method
\begin{align*}
  u^{N+1} = u^{N} - \alpha (dJ^N)^{-1} J(u^N),
\end{align*}
where $dJ^N = -\rho^2 \Delta + f''(u^N)$ is the Fr\'{e}chet derivative
of $F$, and $\alpha$ is the step size which will be determined using
backtracking. As expected, the biggest computational challenge is
solving $- \rho^2 \Delta v + f''(u^{N}) v = J[u^N]$. Solving this
variable coefficient PDE with an integral equation is challenging since
a fundamental solution is not readily available. We propose to instead
use an inexact Newton's method where $dJ^{-1}$ is defined in terms of
the piecewise quadratic approximation
\begin{align*}
  f(u) \approx \tilde{f}(u) = bu + \begin{cases}
    a(u - u_0)^2, &\mbox{if } u \leq (u_0 + u_1)/2, \\
    (u - u_1)^2, &\mbox{if } u > (u_0 + u_1)/2.
  \end{cases}
\end{align*}
and the resulting inexact Newton's method requires solving the piecewise
constant PDE
\begin{align}
  \label{eq:screenedPoisson}
  -\rho^2 \Delta v + \tilde{f}''(u^{N})v = J[u^N],
\end{align}
which is a screened Poisson equation with piecewise a constant screening
coefficient. We test the efficacy of exact and inexact Newton's method
applied to~\eqref{eq:F} in $(0,1)$ by discretizing $dJ$ with a
second-order finite difference method. The choice of parameters $a$,
$b$, $U_0$, and $U_1$ result in the double well potential illustrated in
Figure~\ref{fig:CA}(a), and both the exact and inexact Newton's methods
converge to the solution in Figure~\ref{fig:CA}(b).
Figure~\ref{fig:CA}(c) shows the max norm of the residual at each Newton
iterate when using the exact second derivative (red curve) and the
piecewise constant approximate (blue curve). While the inexact Newton's
method requires roughly twice as many iterations to reach of a tolerance
of $10^{-4}$, each iteration is significantly cheaper in higher
dimensions.

\begin{wrapfigure}[28]{l}{1.8in}
  \vspace{-4pt}
  \centering
  \input{figures/doublewell.tikz} 
  \caption{\label{fig:CA} (a) The double well potential $u_0 =
  0.5$, $u_1 = 2$, $a = 1$, $b = 0.4$. (b) The solution of the
  Allen-Cahn equation~\eqref{eq:SL}. (c) The convergence of the exact
  and inexact Newton's methods.}
\end{wrapfigure}
Having demonstrated that the inexact Newton method converges in one
dimension, it will be coupled with IE formulations to to
solve~\eqref{eq:SL} in two and three dimensions. We first partition
$\Omega$ into $\Omega_0 = \{\xx \in \Omega \:|\: u(\xx) < (u_0 +
u_1)/2\}$ and $\Omega_1 = \{\xx \in \Omega \:|\: u(\xx) > (u_0 +
u_1)/2\}$, and we let $\Lambda = \{\xx \in \Omega \:|\: u(\xx) = (u_0 +
u_1)/2 \}$ be their common boundary, and we let $G_0$ be the fundamental
solution of $(-\rho^2 \Delta - 2a)$ and $G_1$ be the fundamental
solution of $(-\rho^2 \Delta + 2)$. Then, the solution
of~\eqref{eq:screenedPoisson} can be written as $u = u^P + u^H$ where
$u^P$ is a volume potential and $u^H$ is a layer potential. The volume
potential is
\begin{align*}
  u^{P}(\xx) = \int_{\Omega_i} G_i(\xx,\yy) J[u^N](\yy)\, d\yy, 
    \quad i = 1,2,
\end{align*}
but it is discontinuous across $\Lambda$ and does not satisfy the
prescribed hydrophobic boundary conditions on the particles $\Gamma_i$.
Assuming the particles are entirely contained in the hydrophobic region,
the boundary conditions will be enforced with the layer potentials
\begin{align*}
  u^{H}(\xx) = \begin{cases}
    \displaystyle\int_{\Lambda} G_0(\xx,\yy) \eta_0(\yy) \, ds_\yy + 
    \displaystyle\int_{\Gamma} \pderiv{}{\nn_\yy} G_0(\xx,\yy)
      \sigma(\yy) \,ds_\yy, \quad & \xx \in \Omega_0, \\
    \displaystyle\int_{\Lambda} G_1(\xx,\yy) \eta_1(\yy) \, ds_\yy,
     & \xx \in \Omega_1.
  \end{cases}
\end{align*}
The density functions $\sigma$, $\eta_0$, and $\eta_1$ are determined so
that $u = u^P + u^H$ satisfies the hydrophobic boundary condition on
$\Gamma = \cup_i \Gamma_i$, and that it is continuous and differentiable
across $\Lambda$.
